# Class 7: Taxicab problem


## Exercise 1: Taxicabs with JAGS

:::{.example #taxicabs name="Taxicabs with JAGS" .lizi}
A simple implementation of the taxicab problem with a single sample is presented using JAGS. 
:::


```{r}
library(R2jags)

model= function(){
M ~ dcat(p1)
# sampling distribution is uniform over first N integers
# use step function to change p1[j] to 0 for j>N
for (j in 1:1000) {
p1[j] <- step(N - j + 0.01)/N
}
N ~ dcat(p2)
for (j in 1:1000) {
recip[j] <- 1/j
p2[j] <- recip[j]/sum.recip
}
sum.recip <- sum(recip)
}
data=list("M"=100)
fit <- jags(data=data, model=model,parameters.to.save=c("N","M"), n.chain=2, n.iter=100, n.thin=1, n.burn=0, DIC=FALSE)
fit
```


## Exercise 2: Taxi counter

:::{.exercise #taxi-counter name="Taxi counter" .prob}
Suppose taxis were sampled with the following code
```{r}
myrange=sort(rnbinom(2,5,mu=rgamma(2,shape=50,rate=1/200)))
taxis=sample(myrange[1]:myrange[2],10,replace=TRUE)
```
Knowing the data generating mechanism, construct a Bayesian estimate of the total number of taxis (`diff(myrange)+1`) based on the data provided by `taxis`. You can use JAGS or R as you like. 
:::

